ExpressionView

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.ordering.Rnw
EV: reindented vignettes.

The ordering is done in four steps:
\begin{enumerate}
	\item{\code{simplify}: Reduces the dimensions of the problem by excluding duplicate rows. The multiplicity of each slot is taken into account when calculating the score $Q$.}
	\item{\code{prearrange}: Tries to find a good starting point for the optimization by creating a new order, starting from the first row and adding the consecutive rows one after the other, placing them at the optimal position. For smaller samples, the fitness function is used, while for larger samples only the similarity (scalar product) between rows is taken into account.}
	\item{\code{arrange}: Uses the greedy approach to further increase the alignment score.}
	\item{\code{complexify}: Reinstates the original dimensions of the problem by inserting the omitted rows at the newly found position.}
\end{enumerate}

>>>>

The ordering is done in four steps:
\begin{enumerate}
\item \code{simplify}: Reduces the dimensions of the problem by
  excluding duplicate rows. The multiplicity of each slot is taken
  into account when calculating the score $Q$.
\item \code{prearrange}: Tries to find a good starting point for the
  optimization by creating a new order, starting from the first row
  and adding the consecutive rows one after the other, placing them at
  the optimal position. For smaller samples, the fitness function is
  used, while for larger samples only the similarity (scalar product)
  between rows is taken into account.
\item \code{arrange}: Uses the greedy approach to further increase the
  alignment score.
\item \code{complexify}: Reinstates the original dimensions of the
  problem by inserting the omitted rows at the newly found position.
\end{enumerate}

### 2010-04-21 RExpressionView/inst/doc/ExpressionView.tutorial.Rnw
EV: reindented vignettes.

\subsection{External clustering programs}
Since the structure of biclustering results is independent of the applied method, it is straightforward to
import results obtained from external clustering programs and convert them to \Rclass{ISAModules}.
To illustrate the conversion, let us consider the sample data and {\bf randomly} assign the  \Sexpr{nrow(ALL)} genes 
and  \Sexpr{ncol(ALL)} samples to  \Sexpr{length(modules)} modules. The resulting modules can be described by two
binary matrices
<<random-modules>>=
modules.genes <- matrix(as.integer(runif(dim(ALL)[1]*length(modules),0,1)>0.8),nrow=dim(ALL)[1])
modules.conditions <- matrix(as.integer(runif(dim(ALL)[2]*length(modules),0,1)>0.8),nrow=dim(ALL)[2])
@
indicating if a given gene \variable{i} is contained in module \variable{j} if \variable{modules.genes[i,j]$\ne$0}.
Using these matrices, it is straightforward to create an \Rclass{ISAModules} object:
<<toisa>>=
new("ISAModules",
genes=modules.genes, conditions=modules.conditions,rundata=data.frame(),seeddata=data.frame())
@

@@

The result is a list containing various mappings between the original data and the optimal arrangement.
Note that the genes and the samples can be ordered separately. Apart form reordering the 
full gene expression matrix, the algorithm also determines the best arrangement of individual biclusters.
The mapping of the genes and the samples contained in bicluster \variable{i} can be accessed by 
<<gene maps, eval=FALSE>>=
optimalorder$genes[i+1]
optimalorder$samples[i+1]
@
The first elements of the lists contain the optimal ordering of the
complete matrix. By default, the \Rfunction{OrderEV} function
runs for roughly one minute, this might not be sufficient to find an appropriate order for data 
containing many overlapping biclusters. The status of the ordering is stored in 
<<status ordering>>=
optimalorder$status
@
If the status is set to \variable{1}, the algorithm has found the optimal solution. A \variable{0} indicates that the
the calculation could not be terminated within the given timeframe. The \Rfunction{OrderEV} function
accepts two additional parameters to circumvent the problem of partial alignment:
One can start the ordering from a given initial configuration, i.e., the result of a previous arrangement
by defining the \Rargument{initialorder} argument
<<status ordering, eval=FALSE>>=
optimalorderp <- OrderEV(modules, initialorder=optimalorder, maxtime=120)
@
and one can increase the time limit by specifying \Rargument{maxtime}. Note that the time is indicated in seconds
and cannot be  smaller than 1.

@@

The ExpressionView Flash applet can be launched from the R environment:

@@

Video tutorials describing how to use the applet can be found on the ExpressionView \\
\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. The screenshot shown in Fig.~\ref{fig:screenshot} 
and the description below illustrate the main features of the applet:

@@

	\item[a] Opens an ExpressionView data file. Note that before opening a new data file, you should restart the applet, i.e., refresh your browser window.
	\item[b] Exports the current view to a pdf file. The file also includes the title (o) of the gene expression data.
	\item[c] Exports the data of the currently viewed module (=bicluster) to a CSV file, that can be opened as a spreadsheet.
	\item[d] In inspect mode, you can use the mouse to explore the gene expression data. The information about the data under the mouse pointer is shown in the Info Panel (t).
	\item[e, f] Zoom and pan modes allow you to restrict the view to a particular part of the gene expression data.
	\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to auto-zoom onto the modules and {\bf e} to see the whole data. In addition to the simple zoom-in feature, you can also use the mouse to select the rectangular area you want to have a closer look at.

@@

	\item[g, h, i, j] Module highlighting and viewing. It is in general impossible to present mutually overlapping biclusters as single rectangles. They are made up of a collection of rectangles. The ordering algorithm used in the R package realigns the gene expression matrix in a way that maximizes the total area of the largest rectangle in every bicluster. The outlines of these parts are drawn in a slightly brighter color than the background, making them easily recognizable.
	\item[g, h] Modules are highlighted as the user moves the mouse over the gene expression data. The two check boxes allow you to choose between highlighting all the parts of a module (Filling) or alternatively only the largest rectangle (Outline). You can also turn it off completely. For data sets with many modules, it can be helpful to restrict highlighting to Outline.
	\item[i, j] Similar to the highlighting, these two check boxes allow you to show either all the parts of the modules (Filling) or only the largest rectangles (Outline). By shift-clicking one of the check-boxes you can switch between showing only the modules or only the gene expression data.
	\item[k] Sets the visibility of the modules layer. Moving the slider to the left fades out the gene expression data, thus focusing on the Biclusters, while towards the opposite direction, the gene expression data moves to the foreground.

@@

	\item[m] Puts the program in fullscreen mode. Note that due to security reasons, it is impossible to enter text in this mode. On Mac OS X, a bug in Flash player prevents you from exporting data in fullscreen mode.
	\item[n] Opens the ExpressionView website, from where you can download sample files and tutorials.

@@

	\item[p] Modules navigator. The Global tab is always available and shows the complete gene expression data. Additional tabs appear as you open individual modules. To close a module, simply move the mouse over the tab and click the close button that appears. 
	\item[q, r, s] Selected genes, samples and modules. The highlighting reflects the selection in the tables (w). The selection is maintained when switching tabs (p). 

@@

	\item[t] Info panel showing the data associated with the current mouse position. The GO and KEGG list contain the five most significant categories and pathways associated with the modules under the mouse pointer.
	\item[u] Lists the selected genes, samples and modules, together with the intersecting modules.

@@

	\item[w] List navigator. Note that depending on the view (p), the lists only show genes and samples contained in the currently viewed module. Modules can also be opened by double-clicking on the corresponding row. The Experiment tab contains a brief description of the data.
	\item[x] Searches the tables for a given expression and restricts the view to the matching entries. The search function uses Perl-style {\bf regular expressions}. By default, the search functions is applied to the whole table. To restrict it to a particular column, shift-click the corresponding column header.
	\item[z] Select a column header to sort the entries according to that column. Shift-click to restrict the search function to that column.

@@

\section{Using ExpressionView with non-gene expression data}
While ExpressionView is designed to work with gene expression data available in the form of a \Rpackage{Bioconductor} \Rclass{ExpressionSet}, it can also be used to visualize other data. Let us for instance use in-silico data generated by the \Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10 overlapping modules:
<<in-silico>>=
library(ExpressionView)
# generate in-silico data with dimensions m x n containing M overlapping modules
# and add some noise
m <- 50
n <- 500
M <- 10
data <- isa.in.silico(num.rows=m, num.cols=n, num.fact=M, noise=0.1, overlap.row=5)[[1]]
modules <- isa(data)
@

@@

Next, we assign the meta data associated with the rows of the data matrix. In this example we use 5 tags labelled ''row tag``:
<<rowdata annotation>>=
rowdata <- outer(1:nrow(data), 1:5, function(x, y) {
  paste("row description (", x, ", ", y, ")", sep="")
})             
rownames(rowdata) <- rownames(data)
colnames(rowdata) <- paste("row tag", seq_len(ncol(rowdata)))
@
And similarly for the columns, using 10 ``column tags'':
<<coldata annotation>>=
coldata <- outer(1:ncol(data), 1:10, function(x, y) {
  paste("column description (", x, ", ", y, ")", sep="")
})
rownames(coldata) <- colnames(data)
colnames(coldata) <- paste("column tag", seq_len(ncol(coldata)))
@

To finish  the description, we add some general information and merge it with the above tables to get a single named list:
<<in-silico data annotation>>=
description <- list(
experiment=list(
	title="Title", 
	xaxislabel="x-Axis Label",
	yaxislabel="y-Axis Label",
	name="Author", 
	lab="Address", 
	abstract="Abstract", 
	url="URL", 
	annotation="Annotation", 
	organism="Organism"),
coldata=coldata,
rowdata=rowdata
)
@
When dealing with gene expression data, the \variable{xaxislabel} is equal to ''genes`` and the  \variable{yaxislabel} is ''samples``. Finally, we export the data set to an ExpressionView file:
<<in-silico data export, eval=FALSE>>=
ExportEV(modules, data, filename="file.evf", description=description)
@

>>>>

\subsection{External clustering programs}
Since the structure of biclustering results is independent of the
applied method, it is straightforward to import results obtained from
external clustering programs and convert them to
\Rclass{ISAModules}. To illustrate the conversion, let us consider the
sample data and {\bf randomly} assign the  \Sexpr{nrow(ALL)} genes and
\Sexpr{ncol(ALL)} samples to  \Sexpr{length(modules)} modules. The
resulting modules can be described by two binary matrices
<<random-modules>>=
modules.genes <- matrix(as.integer(runif(nrow(ALL) * length(modules)) > 0.8), 
                        nrow=nrow(ALL))
modules.conditions <- matrix(as.integer(runif(ncol(ALL) *
                                        length(modules))>0.8),
                        nrow=ncol(ALL))
@
indicating if a given gene \variable{i} is contained in module
\variable{j} if \variable{modules.genes[i,j]$\ne$0}. Using these
matrices, it is straightforward to create an \Rclass{ISAModules}
object: 
<<toisa>>=
new("ISAModules",
    genes=modules.genes, conditions=modules.conditions,
    rundata=data.frame(), seeddata=data.frame())
@

@@

The result is a list containing various mappings between the original
data and the optimal arrangement. Note that the genes and the samples
can be ordered separately. Apart form reordering the full gene
expression matrix, the algorithm also determines the best arrangement
of individual biclusters. The mapping of the genes and the samples
contained in bicluster \variable{i} can be accessed by
<<gene maps, eval=FALSE>>=
optimalorder$genes[i+1]
optimalorder$samples[i+1]
@
The first elements of the lists contain the optimal ordering of the
complete matrix. By default, the \Rfunction{OrderEV} function
runs for roughly one minute, this might not be sufficient to find an
appropriate order for data containing many overlapping biclusters. The
status of the ordering is stored in
<<status ordering>>=
optimalorder$status
@
% $
If the status is set to \variable{1}, the algorithm has found the
optimal solution. A \variable{0} indicates that the the calculation
could not be terminated within the given timeframe. The
\Rfunction{OrderEV} function accepts two additional parameters to
circumvent the problem of partial alignment: One can start the
ordering from a given initial configuration, i.e., the result of a
previous arrangement by defining the \Rargument{initialorder} argument
<<status ordering, eval=FALSE>>=
optimalorderp <- OrderEV(modules, initialorder=optimalorder, maxtime=120)
@
and one can increase the time limit by specifying
\Rargument{maxtime}. Note that the time is indicated in seconds and
cannot be  smaller than 1.

@@

The ExpressionView Flash applet can be launched from the R
environment:

@@

Video tutorials describing how to use the applet can be found on the
ExpressionView website at
\url{http://www.unil.ch/cbg/ExpressionView}. The screenshot shown in
Fig.~\ref{fig:screenshot} and the description below illustrate the
main features of the applet:

@@

\item[a] Opens an ExpressionView data file. Note that before opening a
  new data file, you should restart the applet, i.e., refresh your
  browser window.
\item[b] Exports the current view to a pdf file. The file also
  includes the title (o) of the gene expression data.
\item[c] Exports the data of the currently viewed module (=bicluster)
  to a CSV file, that can be opened as a spreadsheet.
\item[d] In inspect mode, you can use the mouse to explore the gene
  expression data. The information about the data under the mouse
  pointer is shown in the Info Panel (t).
\item[e, f] Zoom and pan modes allow you to restrict the view to a
  particular part of the gene expression data.
\item[e] In zoom mode, you can also use keyboard shortcuts: {\bf a} to
  auto-zoom onto the modules and {\bf e} to see the whole data. In
  addition to the simple zoom-in feature, you can also use the mouse
  to select the rectangular area you want to have a closer look at.

@@

\item[g, h, i, j] Module highlighting and viewing. It is in general
  impossible to present mutually overlapping biclusters as single
  rectangles. They are made up of a collection of rectangles. The
  ordering algorithm used in the R package realigns the gene
  expression matrix in a way that maximizes the total area of the
  largest rectangle in every bicluster. The outlines of these parts
  are drawn in a slightly brighter color than the background, making
  them easily recognizable.
\item[g, h] Modules are highlighted as the user moves the mouse over
  the gene expression data. The two check boxes allow you to choose
  between highlighting all the parts of a module (Filling) or
  alternatively only the largest rectangle (Outline). You can also
  turn it off completely. For data sets with many modules, it can be
  helpful to restrict highlighting to Outline.
\item[i, j] Similar to the highlighting, these two check boxes allow
  you to show either all the parts of the modules (Filling) or only
  the largest rectangles (Outline). By shift-clicking one of the
  check-boxes you can switch between showing only the modules or only
  the gene expression data.
\item[k] Sets the visibility of the modules layer. Moving the slider
  to the left fades out the gene expression data, thus focusing on the
  Biclusters, while towards the opposite direction, the gene
  expression data moves to the foreground.

@@

\item[m] Puts the program in fullscreen mode. Note that due to
  security reasons, it is impossible to enter text in this mode. On
  Mac OS X, a bug in Flash player prevents you from exporting data in
  fullscreen mode.
\item[n] Opens the ExpressionView website, from where you can download
  sample files and tutorials.

@@

\item[p] Modules navigator. The Global tab is always available and
  shows the complete gene expression data. Additional tabs appear as
  you open individual modules. To close a module, simply move the
  mouse over the tab and click the close button that appears. 
\item[q, r, s] Selected genes, samples and modules. The highlighting
  reflects the selection in the tables (w). The selection is
  maintained when switching tabs (p).

@@

\item[t] Info panel showing the data associated with the current mouse
  position. The GO and KEGG list contain the five most significant
  categories and pathways associated with the modules under the mouse
  pointer.
\item[u] Lists the selected genes, samples and modules, together with
  the intersecting modules.

@@

\item[w] List navigator. Note that depending on the view (p), the
  lists only show genes and samples contained in the currently viewed
  module. Modules can also be opened by double-clicking on the
  corresponding row. The Experiment tab contains a brief description
  of the data.
\item[x] Searches the tables for a given expression and restricts the
  view to the matching entries. The search function uses Perl-style
  {\bf regular expressions}. By default, the search functions is
  applied to the whole table. To restrict it to a particular column,
  shift-click the corresponding column header.
\item[z] Select a column header to sort the entries according to that
  column. Shift-click to restrict the search function to that column.

@@

\section{Using ExpressionView with non-gene expression data}
While ExpressionView is designed to work with gene expression data
available in the form of a \Rpackage{Bioconductor}
\Rclass{ExpressionSet}, it can also be used to visualize other
data. Let us for instance use in-silico data generated by the
\Rpackage{isa} package with dimensions 50 $\times$ 500 containing 10
overlapping modules:
<<in-silico>>=
library(ExpressionView)
# generate in-silico data with dimensions m x n 
# containing M overlapping modules
# and add some noise
m <- 50
n <- 500
M <- 10
data <- isa.in.silico(num.rows=m, num.cols=n, 
                      num.fact=M, noise=0.1, overlap.row=5)[[1]]
modules <- isa(data)
@

@@

Next, we assign the meta data associated with the rows of the data
matrix. In this example we use 5 tags labelled ``row tag'':
<<rowdata annotation>>=
rowdata <- outer(1:nrow(data), 1:5, function(x, y) {
  paste("row description (", x, ", ", y, ")", sep="")
})             
rownames(rowdata) <- rownames(data)
colnames(rowdata) <- paste("row tag", seq_len(ncol(rowdata)))
@
And similarly for the columns, using 10 ``column tags'':
<<coldata annotation>>=
coldata <- outer(1:ncol(data), 1:10, function(x, y) {
  paste("column description (", x, ", ", y, ")", sep="")
})
rownames(coldata) <- colnames(data)
colnames(coldata) <- paste("column tag", seq_len(ncol(coldata)))
@

To finish  the description, we add some general information and merge
it with the above tables to get a single named list:
<<in-silico data annotation>>=
description <- list(
experiment=list(
	title="Title", 
	xaxislabel="x-Axis Label",
	yaxislabel="y-Axis Label",
	name="Author", 
	lab="Address", 
	abstract="Abstract", 
	url="URL", 
	annotation="Annotation", 
	organism="Organism"),
coldata=coldata,
rowdata=rowdata
)
@
When dealing with gene expression data, the \variable{xaxislabel} is
equal to ``genes'' and the  \variable{yaxislabel} is
``samples''. Finally, we export the data set to an ExpressionView
file: 
<<in-silico data export, eval=FALSE>>=
ExportEV(modules, data, filename="file.evf", 
         description=description)
@
